A best practices composite lifetable for US states

### 2015-09-24 BPLE/Manuscript.tex
added in most of Magali's changes, now to look at Adrien's email

\FloatBarrier
\section*{Preliminary results}
\FloatBarrier

@@

@@

We decompose the state departures in life expectancy from the best practice
trend in each year and sex, over both age and cause. Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always
of the same sign, which makes interpretation of the heatmaps more intuitive.
Deeper shades of red indicate larger departures, i.e., larger $e(0)$
shortcomings.
In Figure~\ref{fig:dep} we represent the average $e(0)$ shortcoming from each
cause and state, and average within decades to summarize the total contribution
to the departure from best practices.\footnote{The 2000s trend is based on the average for the years 2000-2004 only.} Major causes of death are in
rows, and decades are organized in columns.

In these heat maps, states are represented as identically-sized square cells,
while preserving rough spatial relationships from a standard map
projection.\footnote{This configuration is inspired by a tile grid map on the
NPR blog \citep{NPRsquares}, which
itself was inspired by various commonly used maps in the New York Times,
Washington Post, and others. DeBelius notes that grid panels are not meant for
displaying population quantities, since each state has a different population
size. Since lifetable quantities are purged of population size and
structure, the visual instrument is adequate, and the equalization of state
size and form adds legibility to our results.} For orientation, we present a
state guide to the heatmaps in Figure~\ref{fig:heatguide}.
\begin{figure}
\centering
\caption{A guide to tiled heatmaps of U.S. states.}
\label{fig:heatguide}
\includegraphics[scale=.3]{Figures/StatesDiagram.pdf}
\end{figure}

For the case of cardiovascular disease
and other causes, grouping departures into the 2$+$ group blends out much
interstate variation, since some contributions range as high as 4
years, but broad patterns are still visible.\footnote{This visual display may be
significantly adapted when we add more causes to the decomposition. We will also do a similar small multiples
display for grouped age patterns.}

Since the cause groupings available at the time of submission are not the final groupings, and
because the causes included in the study contribute such varying magnitudes to
state differences, trends are dominated by Cardivascular diseases and the large
aggregate group of all other causes. The 11 original causes were grouped into
five so as to make state patterns more visible. More commentary to come at a
later date.

@@

\section*{}

\FloatBarrier

>>>>

There are limitations to the interpretation of a best practices trend, as
defined here. The mix of causes within each age, and the resulting age pattern
for each cause may not necessarily reflect a possible or likely mortality
pattern. For instance, diseases may compensate each other, especially for
diagnostic and reporting reasons. For instance, for decedents that suffered
from both condition A and condition B, one state might be more likely to
declare A as the underlying cause of death, while another state would more
likely declare B to be as the underlying cause. A situation like this
would tend to inflate the best practices $e(0)$. The best practices $e(0)$ may
be underestimate in instances where reductions in cause A are expected to lead
to reductions in cause B (e.g., pneumonia and cardiovascular disease.)
\FloatBarrier
\section*{Preliminary results}
\FloatBarrier

@@

For the sake of comparison with the presentlt calculated best practices series.
The $e(0)$ value in 2004 for males and females was 80 and 84.9, respectively. In
the same year, the observed all-cause $e(0)$ in Japan were 78.6 for males and
85.5 for females. That is, the U.S. best practices trend is ahead of observed
life expectancy for Japanese males, and behind for Japanese females. We may also
compare with the official $e(0)$ projections of the U.S. Social Security
Administration (SSA). The SSA projects projects male $e(0)$ passing 80 in the
year 2058, and female $e(0)$
passing 84.9 in the year 2075 \citep{SSA2005}. These projections are a full 54
and 71 years after 2004, in both cases much farther in the future than a simple extrapolation
of the best practices trend would predict, circa 2034.

@@

Since the trend in the best practices $e(0)$ is so stable over time, it
is also a convenient moving benchmark against which to compare state patterns
--- a kind of potential mortality for each state. We
decompose the state departures in life expectancy from the best practice trend in each year and sex, over both age and cause. Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always of the same sign, which makes heatmaps more intuitive than would otherwise be the case.\footnote{Typically age and/or cause-specific contributions to a difference in $e(0)$ are represented in stacked bar charts, which in the present case would not be a practical way to distill information.} We therefore summarize state trends in a series of heatmaps representing the 50 states plus the District of Columbia. In these heat maps, states are represented as identically-sized square cells, while preserving rough spatial relationships from a standard map projection. For orientation, we present a state guide to the heatmaps in Figure~\ref{fig:heatguide}.\footnote{This configuration is inspired by a tile grid map on the
NPR blog \citep{NPRsquares}, which
itself was inspired by various commonly used maps in the New York Times,
Washington Post, and others. DeBelius notes that grid panels are not meant for
displaying population quantities, since each state has a different population
size. Since lifetable quantities are purged of population size and
structure, the visual instrument is adequate, and the equalization of state
size and form adds legibility to our results.} Deeper shades of red
indicate larger departures, i.e., larger $e(0)$ shortcomings.

\begin{figure}
\centering
\caption{A guide to tiled heatmaps of U.S. states.}
\label{fig:heatguide}
\includegraphics[scale=.3]{Figures/StatesDiagram.pdf}
\end{figure}

In Figures~\ref{fig:depm} and \ref{fig:depf} we represent the average $e(0)$
shortcoming from each cause and state, and average within decades to summarize the total contribution
to the departure from best practices.\footnote{The 2000s results are based on
the average for the years 2000-2004 only.} Major causes of death are in rows,
and decades are organized in columns. For the case of cardiovascular disease and other causes, grouping departures into the 2$+$ group blends out much
interstate variation, since some contributions range as high as 4
years, but broad patterns are still visible.\footnote{This visual display may be
significantly adapted when we add more causes to the decomposition. We will also do a similar small multiples
display for grouped age patterns.}

Glancing over the lung diseases and cancer rows of the maps, it is evident that
the District of Columbia had high excess mortality for these causes over all
five decades, peaking in the 1980s. Lung diseases display a clear mortality
basin in the Southern states for males, although the absolute maginitude of these departures from ideal mortality has diminished over time. The same mortality basin occurs in
cariovascular causes over all five decades, but is clearly visible in this
display only since the 1990's, due to the 2$+$ grouping.

Since the cause groupings available at the time of submission are not the final groupings, and
because the causes included in the study contribute such varying magnitudes to
state differences, trends are dominated by Cardivascular diseases and the large
aggregate group of all other causes. The 11 original causes were grouped into
five so as to make state patterns more visible. These results will be commented
at greater length when the final results are prepared.

@@

\FloatBarrier
\section*{Discussion}
In this paper, we have given an intuitive rationale for calculating a best
practices trend in the United States. First, the best practices trend in life
expectancy at birth can be interpreted as the mortality potential present in the
United States at a given point in time. It is an indication of the mortality
conditions that would come to be if local conditions, practices,
technologies, and behaviors were perfectly diffused from those places with the
best age and cause-specific performance in the year. While it is difficult to
imagine that pressing a magic button to activate this perfect cocktail of
mortality risk factors would actually bring about this minimum mortality
instantaneously--- for instance, cumulative risk factors in population
stocks that are distributed unevenly over space would stymie such a thought
experiment--- we do observe catch-ups with past best practices values for many states, and for the
U.S. national $e(0)$, on average after a period of 30 years. It remains to be
seen how this lag scales to the number of causes ocnsidered and size and
distribution of subpopulations. At this time, we are rather attracted to the
idea of a period weathervane that points us to where period mortality is going.
